---
title: "Model fitting in R"
author: "Courtney Goodridge"
date: "09/05/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r}
if(!require(here)) install.packages("here")
library(here)

if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

if(!require(dplyr)) install.packages("dplyr")
library(dplyr)

if(!require(tidyr)) install.packages("tidyr")
library(tidyr)

if(!require(viridis)) install.packages("viridis")
library(viridis)

if(!require(lme4)) install.packages("lme4")
library(lme4)

if(!require(data.table)) install.packages("data.table")
library(data.table)

if(!require(emmeans)) install.packages("emmeans")
library(emmeans)

if(!require(patchwork)) install.packages("patchwork")
library(patchwork)

if(!require(marginaleffects)) install.packages("marginaleffects")
library(marginaleffects)

if(!require(purrr)) install.packages("purrr")
library(purrr)

if(!require(scales)) install.packages("scales")
library(scales)

if(!require(afex)) install.packages("afex")
library(afex)

if(!require(effectsize)) install.packages("effectsize")
library(effectsize)
```

### Study 1: Mole, C., Pekkanen, J., Sheppard, W., Louw, T., Romano, R., Merat, N., ... & Wilkie, R. (2020). Predicting takeover response to silent automated vehicle failures. PLoS One, 15(11), e0242825.

## Fitting an ordinary linear regression

```{r}
# fitting a linear regression model
mod_1 <- lm(TLC_takeover ~ TLC_failure, data = data_RTs_cens_NA_removed)

# summarizing the linear regression model to see the parameter estimates
summary(mod_1)

# confidence intervals for the linear regression
confint(mod_1)
```

## Fitting a varying intercept model

Here we fit a varying intercept model. The *(1 | ppid)* part of the model equations fits an intercept for each participant. This accounts for the non-independence of observations by constraining the non-independent clusters to the same intercept. Remember, this is something we need to do as each member of the random effects group has multiple observations. 

```{r}
# fitting a varying intercept model
mod_2 <- lme4::lmer(TLC_takeover ~ TLC_failure + (1 | ppid), data = data_RTs_cens_NA_removed)

# summarizing the model
summary(mod_2)

# generating confidence intervals
confint(mod_2)

coef(mod_2)[1]
```

## Fitting a varying intercept, varying slope model

Here we fit a varying intercept, varying slope model. The *(TLC_failure | ppid)* part of the model equations fits an intercept for each participant, and allows each participants slope to vary. This accounts for the non-independence of observations by constraining the non-independent clusters to the same intercept, and models the differences in sensitivity between different people in the sample. 

```{r}
# fitting a varying intercept. varying slope model
mod_3 <- lme4::lmer(TLC_takeover ~ TLC_failure + (TLC_failure | ppid), data = data_RTs_cens_NA_removed)

# summarizing the model
summary(mod_3)

# generating confidence intervals
confint(mod_3)

coef(mod_3)[1]
```

### Study 2: Goodridge, C. M., Goncalves, R. C., Arabian, A., Horrobin, A., Solernou, A., Lee, Y. T., ... & Merat, N. (2024). Gaze entropy metrics for mental workload estimation are heterogenous during hands-off level 2 automation. Accident Analysis & Prevention, 202, 107560.

## Model fitting

```{r}
mod_4 <- lmer(e.norm ~ n_back * lead + (n_back | ppid), data = goodridge_2023_dat)

summary(mod_4)

confint(mod_4)
```


## Maximal models

```{r}
mod_5 <- lmer(e.norm ~ n_back * lead + (n_back * lead | ppid), data = goodridge_2023_dat)

summary(mod_5)
```

## Removing correlation parameters

```{r}


```


## Comparing MLM outputs to RM ANOVAs

```{r}
rm_anova <- aov_car(e.norm ~ n_back * lead + Error(ppid/n_back * lead), data = goodridge_2023_dat)

summary(rm_anova)
eta_squared(rm_anova)

goodridge_2023_dat %>%
  dplyr::group_by(n_back) %>%
  dplyr::summarise(mean_e = mean(e.norm), mean_sd = sd(e.norm))
```






