---
title: "Multilevel Models - Applications to Human Factors Research"
author: "Courtney Goodridge"
date: "17/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script contains the analysis for the manuscripts: *Multilevel Models: Applications and Uses in Human Factors Research*

Data for the first analysis is from the Mole et al (2020) manuscript. This manuscript reference is: Mole, C., Pekkanen, J., Sheppard, W., Louw, T., Romano, R., Merat, N., ... & Wilkie, R. (2020). Predicting takeover response to silent automated vehicle failures. Plos one, 15(11), e0242825. 

Data and analysis scripts can be found on OSF: https://osf.io/aw8kp/. 


First we load the necessary packages needed for this analysis

## Packages

```{r}
if(!require(here)) install.packages("here")
library(here)

if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

if(!require(dplyr)) install.packages("dplyr")
library(dplyr)

if(!require(tidyr)) install.packages("tidyr")
library(tidyr)

if(!require(viridis)) install.packages("viridis")
library(viridis)

if(!require(lme4)) install.packages("lme4")
library(lme4)

if(!require(data.table)) install.packages("data.table")
library(data.table)
```

## Data

```{r}
steering_data <- fread(file = here::here("ITS/Multilevel models paper/Mole et al (2020) data and analysis/collated_steering.csv"))
```

## Calculating reaction time

Now the data has been loaded, the reaction time can be calculated. I calculate steering wheel angle and mirror the directional variables for completeness but this is not strictly necessary (only TLC_takeover will be used for the analysis). 

```{r}
# Mole et al (2020) calculating the SWA. I won't be using that for the analysis, but for completeness I calculate it here.
steering_data <- steering_data %>%
  dplyr::rename(swv = swa) %>%
  dplyr::mutate(swa = swv * 90)

# Next the data mirrored. Again, I am only focusing on timing variable so these won't be needed for the analysis but are included for completeness.
steering_data <- steering_data %>% 
  dplyr::mutate(world_x_mirrored = if_else(bend == -1, world_x * -1, world_x),
                swa_mirrored = if_else(bend == -1, swa * -1, swa),
                swv_mirrored = if_else(bend == -1, swv * -1, swv),
                sb_mirrored = if_else(bend == -1, steeringbias * -1, steeringbias))

# function for calculating the TLC from the offset where is varies 
TLC_from_offset <- function(b, w = 1.5, r = 80, v = 8){
  b = b / 180 * pi
  
  TLC = sqrt(w*(2*r + sign(b)*w)/(abs(b)*r*v))
}

steering_data <- steering_data %>% 
  dplyr::mutate(TLC_calc = TLC_from_offset(sab))


#add RT and disengage flag. This code calculates RT. I had to alter it slightly to include an if statement where if the auto_false was length 0 (i.e., a person never disengaged), it was set to NA instead of being empty. This allows the is.na() function to work in the following lines of code)
disengage_RT <- function(onsettime, timestamp_trial, autoflag){
  
  #pick first frame where autoflag == false, then take the timestamp and minus the onset_time
  auto_false <- which(autoflag == FALSE) # 1) select the autoflags which are false
  if(length(auto_false) == 0){ # 2) if there is no false autoflag, set auto_false to NA
    auto_false = NA
  }
  disengage_index <- first(auto_false) # 3 select the first frame that automation was disengaged
  disengage_trialtime <- timestamp_trial[disengage_index] # 4) select the trial time of the automation disengagement
  onset_time <- first(onsettime) # 5 Select the first onset time
  RT <- disengage_trialtime - onset_time # 6 subtract onset time from disengage trial time (can be negative)
  return(RT) # return the RT
  
}

# calculate RT
steering_data <- steering_data  %>% 
  dplyr::group_by(ppid, sab, cogload, trialn) %>% 
  dplyr::mutate(RT = disengage_RT(onsettime, timestamp_trial, autoflag),
         disengaged = ifelse(is.na(RT), 0, 1) # whether or not they actually took over.
  )

# change cogload variable into a factor
steering_data$cogload <- as.factor(steering_data$cogload)


#rename cogload factors so that it makes sense - no load and load
steering_data$cogload <- plyr::mapvalues(steering_data$cogload, from = c("None", "Middle"), to = c("noload", "load"))

#create unique trial id
steering_data <- steering_data %>% 
  dplyr::mutate(trialid = paste(ppid, cogload, trialn, sep = "_"))

# select reaction times for each condition
data_RTs <- steering_data %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(ppid, cogload, trialn) %>% 
  dplyr::summarise(RT = first(RT),
                   disengaged = first(disengaged), #whether or not they actually took over.
                   premature = ifelse(RT <= 0, 1, 0),
                   sab = first(sab),
                   maxsb = max(abs(sb_mirrored)),
                   onsettime = first(onsettime),
                   design = first(design),
                   #simTLC = first(simulated_TLC),
                   autofile_i = first(autofile_i),
                   TLC_calc = first(TLC_calc))
```

## Calculating TLC_takeover

To calculate the TLC_takeover, the simulated timings for the silent failures need to be merged with the reaction times. This is done in the following chunk. 

```{r}
# loading simualted_TLC
sim_TLCs <- fread(file = here::here("ITS/Multilevel models paper/Mole et al (2020) data and analysis/simulated_ttlcs.csv"))

# merge simulated TLCs with RT dataframe
data_RTs <- merge(data_RTs, sim_TLCs, by = c("sab","onsettime","autofile_i"))

# calculating TLC_takeover
data_RTs <- data_RTs %>% 
  dplyr::mutate(TLC_onset = simulated_ttlc,
                TLC_takeover = TLC_onset - RT,
                time_remaining = 15 - onsettime,
                early = time_remaining - RT,
                TLC_end_of_trial = TLC_onset - time_remaining)  

# filter out conditions where takeovers did not happen, or they happened too early (i.e., before the onset of the failure)
data_RTs_cens <- data_RTs %>%
  dplyr::filter(premature == 0 | is.na(premature)) %>% 
  dplyr::mutate(censored = ifelse(is.na(TLC_takeover), -1,
                                  ifelse(TLC_takeover < TLC_end_of_trial, -1, 0)))

data_RTs_cens$cogload <- relevel(data_RTs_cens$cogload, "noload")

#13 moves the wheel pretty much every trial before they should do so they are filtered out
data_RTs_cens <- dplyr::filter(data_RTs_cens, ppid != 13) 

# removing NAs (i.e, when people did not disengage) and only concentrating on no load conditions
data_RTs_cens_NA_removed <- data_RTs_cens %>%
  dplyr::filter(disengaged == 1, cogload == "noload")

# only focusing on balanced conditions and rounding the onset values
data_RTs_cens_NA_removed <- data_RTs_cens_NA_removed %>% 
  dplyr::filter(design == "balanced") %>% 
  dplyr::mutate(TLC_failure = round(TLC_onset, digits = 2)) 
```

