---
title: 'Multilevel Models - Applications to Human Factors Research: Read me plots'
author: "Courtney Goodridge"
date: "09/01/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r}
if(!require(here)) install.packages("here")
library(here)

if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

if(!require(dplyr)) install.packages("dplyr")
library(dplyr)

if(!require(tidyr)) install.packages("tidyr")
library(tidyr)

if(!require(viridis)) install.packages("viridis")
library(viridis)

if(!require(lme4)) install.packages("lme4")
library(lme4)

if(!require(data.table)) install.packages("data.table")
library(data.table)

if(!require(emmeans)) install.packages("emmeans")
library(emmeans)

if(!require(patchwork)) install.packages("patchwork")
library(patchwork)

if(!require(marginaleffects)) install.packages("marginaleffects")
library(marginaleffects)

if(!require(purrr)) install.packages("purrr")
library(purrr)
```

## Data

Load data for read me file plots

```{r}




```

## General assumption

Lets assume with have a variable X which we think might be related variable Y

```{r}
fig_1_readme <- ggplot() +
  geom_point(data_RTs_cens_NA_removed %>%
               dplyr::filter(ppid <= 6), mapping = aes(x = TLC_failure, TLC_takeover), position = position_jitter(seed = 42, width = 2), size = 1.5) +
  xlab("X") +
  ylab("Y") +
  ylim(0, 7) +
  xlim(0, 11) +
  theme_plot

ggsave(here::here("multilevel_model_HF_applications/readme_plots/fig_1_readme.tiff"), plot = fig_1_readme, width = 6, height = 6, units = 'cm', dpi = 300, type = 'cairo')
```

## Traditional analysis 

A traditional analysis might fit a model to predict Y as a function of X. 

```{r}
readme_mod_1 <- lm(TLC_takeover ~ TLC_failure, data = data_RTs_cens_NA_removed %>%
                    dplyr::filter(ppid <= 6))

readme_mod_emmeans <- emmeans(object = readme_mod_1, specs = ~ TLC_failure, at = list(TLC_failure = seq(0, to = 12, by = .1))) %>% 
  data.frame()

fig_2_readme <- ggplot() +
  geom_point(data_RTs_cens_NA_removed %>%
               dplyr::filter(ppid <= 6), mapping = aes(x = TLC_failure, TLC_takeover), position = position_jitter(seed = 42, width = 2), size = 1.5) +
  geom_ribbon(data = readme_mod_emmeans, mapping = aes(x = TLC_failure, y = emmean, ymin = lower.CL, ymax = upper.CL), alpha = .5) +
  geom_line(readme_mod_emmeans, mapping = aes(x = TLC_failure, y = emmean), linetype = "dashed", col = "red") +
  xlab("X") +
  ylab("Y") +
  ylim(0, 7) +
  xlim(0, 11) +
  theme_plot

ggsave(here::here("multilevel_model_HF_applications/readme_plots/fig_2_readme.tiff"), plot = fig_2_readme, width = 6, height = 6, units = 'cm', dpi = 300, type = 'cairo')
```

## It's a complicated world

Not everyone or everything is the same. There are groups within the whole that might differ in interesting ways. Perhaps for some of those groups, X predicts Y very strongly; for others, less so.

```{r}
fig_3_readme <- ggplot() +
  geom_point(data_RTs_cens_NA_removed %>%
               dplyr::filter(ppid <= 6), mapping = aes(x = TLC_failure, TLC_takeover, col = as.factor(ppid)), position = position_jitter(seed = 42, width = 2), size = 1.5) +
  geom_ribbon(data = readme_mod_emmeans, mapping = aes(x = TLC_failure, y = emmean, ymin = lower.CL, ymax = upper.CL), alpha = .5) +
  geom_line(readme_mod_emmeans, mapping = aes(x = TLC_failure, y = emmean), linetype = "dashed", col = "red") +
  xlab("X") +
  ylab("Y") +
  ylim(0, 7) +
  xlim(0, 11) +
  theme_plot +
  theme(legend.position = "none")

ggsave(here::here("multilevel_model_HF_applications/readme_plots/fig_3_readme.tiff"), plot = fig_3_readme, width = 6, height = 6, units = 'cm', dpi = 300, type = 'cairo')
```

## Multilevels models can help make sense

```{r}
readme_mod_2 <- lmer(TLC_takeover ~ TLC_failure + (TLC_failure | ppid), data = data_RTs_cens_NA_removed %>%
                    dplyr::filter(ppid <= 6))

example <- data_RTs_cens_NA_removed %>%
  dplyr::filter(ppid <= 6)
  
# generate new data
failures <- data.frame(TLC_failure = seq(0, to = 12, by = .1))
ppids <- data.frame(ppid = unique(example$ppid))
new_values <- expand_grid(failures, ppids)

# predict ppid-level estimates from the multilevel model
readme_intercepts_slopes <- predict(readme_mod_2, newdata = new_values) %>% 
  data.frame() %>% 
  set_names("y_hat") %>% 
  dplyr::bind_cols(new_values)

readme_intercepts_slopes_emmeans <- emmeans(object = readme_mod_2, specs = ~ TLC_failure, at = list(TLC_failure = seq(0, to = 12, by = .1))) %>% 
  data.frame()


fig_4_readme <- ggplot() +
  geom_point(data_RTs_cens_NA_removed %>%
         dplyr::filter(ppid <= 6), mapping = aes(x = TLC_failure, TLC_takeover, col = as.factor(ppid)), position = position_jitter(seed = 42, width = 2),size = 1.5, alpha = .5) +
  geom_line(readme_intercepts_slopes, mapping = aes(x = TLC_failure, y = y_hat, col = as.factor(ppid))) +
  xlab("X") +
  ylab("Y") +
  ylim(0, 7) +
  xlim(0, 11) +
  theme_plot +
  theme(legend.position = "none")

# faceting by group - not using this plot in the read me though
ggplot() +
  geom_point(data_RTs_cens_NA_removed %>%
         dplyr::filter(ppid <= 6), mapping = aes(x = TLC_failure, TLC_takeover, col = as.factor(ppid)), position = position_jitter(seed = 42, width = 2),size = 1.5) +
  geom_line(readme_intercepts_slopes, mapping = aes(x = TLC_failure, y = y_hat, col = as.factor(ppid))) +
  xlab("X") +
  ylab("Y") +
  ylim(0, 7) +
  xlim(0, 11) +
  theme_plot +
  facet_wrap(~ ppid) +
  theme(legend.position = "none")

ggsave(here::here("multilevel_model_HF_applications/readme_plots/fig_4_readme.tiff"), plot = fig_4_readme, width = 6, height = 6, units = 'cm', dpi = 300, type = 'cairo')
```